---
title: "Benchmarking computational efficiency of TreeSE methods"
author: "Giulio Benedetti and Leo Lahti"
date: "`r Sys.Date()`"
---


## Overview

This report contains benchmarking results for the `r testmethod`
operation with common microbiome data containers in R/Bioconductor.

The benchmarking tests utilize the following publicly available data
sets. We thank the original authors for making these valuable data
resources openly available. Check the links for details and original
references:

- [hitchip1006](https://github.com/microbiome/miaTime/blob/master/R/data.R) (Lahti et al. 2014)
- [SongQAData](https://microbiome.github.io/microbiomeDataSets/reference/SongQAData.html) Song et al. (2016)
- [GrieneisenTSData](https://microbiome.github.io/microbiomeDataSets/reference/GrieneisenTSData.html) Grieneisen et al. (2021) baboon data set



```{r setup, echo = FALSE}
library(Cairo)

# global knitr options
knitr::opts_chunk$set(fig.path='figure/',
                      fig.prefix=testmethod,
                      dpi=300,
                      dev = "png",
                      dev.args = list(type = "cairo-png"),
		      # fig.width=10		      
		      )
```


## Data characteristics


Full sample sizes by data set:

```{r samples, echo = FALSE, fig.width = 10, fig.height = 4*length(containers)}
samples <- data.frame(Dataset=sapply(containers, mainExpName), N=sapply(containers, ncol)) %>% arrange(N)
kable(samples)
```

Feature counts by data set:

```{r chars, echo = FALSE, fig.width = 10, fig.height = 4*length(containers)}
feat <- df_benchmark %>% select(Dataset, Rank, Features) %>% unique() %>%
                    pivot_wider(names_from=Dataset, values_from=Features, values_fill=0)
feat <- feat[order(feat[,2]), ] # Sort by the first dataset feature counts
kable(feat)
```


## Relative differences in execution time by sample size

```{r ratio, echo = FALSE, fig.height = 5, fig.width = 4*length(containers)}
df <- df_benchmark %>% select(Dataset, Time, Features, Samples, Rank, ObjectType) %>% 
      pivot_wider(names_from = c(ObjectType), values_from = Time) %>% 
      mutate(Ratio = tse / pseq) 

p <- ggplot(df, aes(x = Samples, y = Ratio, group = Rank, color = Features)) + 
    geom_point() + 
    geom_line() +
    labs(title = "Relative difference in execution times",
         x = "Samples (N)",
         y = "Relative time (TreeSE / phyloseq)",
         color = "Features (N)",
         caption = "Execution time ratio by samples (TreeSE/phyloseq)") +
    # scale_y_continuous(label = scales::percent) +
    scale_x_log10() +
    facet_grid(. ~ Dataset) + 
    geom_hline(yintercept = 1, linetype = 2) 

print(p)
```



## Absolute execution time by sample size

```{r abs_by_time, echo = FALSE, fig.height = 13, fig.width = 3*length(containers)}
df <- df_benchmark %>% select(Dataset, Time, Features, Samples, Rank, ObjectType)
df <- df %>% filter(Dataset %in% names(which(table(df$Dataset) >= 10)))
cols <- c(tse="black",pseq="darkgray") # Ensure colors always in same order

p <- ggplot(df, aes(x = Samples, y = Time * 1000, color = ObjectType)) + 
    geom_point() + 
    geom_line() +
    theme(axis.text.x=element_text(angle=90)) + 
    labs(title = "Absolute times",
         x = "Samples (N)",
         y = "Time (ms)",
         color = "Container",
         caption = "Execution time comparison between TreeSE and phyloseq."
	 ) +
	 scale_color_manual(values=cols[unique(df$ObjectType)]) + 
    facet_grid(Rank ~ Dataset, scales = "free_y") 

print(p)
```


## Execution times vs number of features

```{r multi_ex_time, echo = FALSE, fig.height=13, fig.width = 3*length(containers)}
p <- ggplot(df_benchmark, aes(x = Features, y = Time * 1000, color = ObjectType)) + 
    geom_point() + 
    geom_line() +
    labs(title = "Feature count vs. execution time",
         x = "Features (D)",
         y = "Execution time (ms)",
         color = "Method:",
         caption = "Execution time as a function of number of features") +
    scale_x_log10() + # Log is often useful with sample size
    # scale_y_log10() + # Log is often useful with sample size
    scale_color_manual(values=cols[unique(df$ObjectType)]) +     
    # facet_grid(Rank ~ Samples, scales = "free_y")
    facet_grid(Samples ~ Dataset, scales = "free_y")    

print(p)
```



