---
title: "`r testmethod`"
subtitle: "Benchmarking computational efficiency of microbiome data containers"
author: "Giulio Benedetti and Leo Lahti"
date: "`r Sys.Date()`"
---


## Overview

This report contains benchmarking results for the `r testmethod`
operation with common microbiome data containers in R/Bioconductor for
the `r testmethod` operation.

The tests utilize the following publicly available data sets. We thank
the original authors for making these valuable data resources openly
available. Check the links for details and original references:

- [hitchip1006](https://github.com/microbiome/miaTime/blob/master/R/data.R) (Lahti et al. 2014)
- [SongQAData](https://microbiome.github.io/microbiomeDataSets/reference/SongQAData.html) Song et al. (2016)
- [GrieneisenTSData](https://microbiome.github.io/microbiomeDataSets/reference/GrieneisenTSData.html) Grieneisen et al. (2021) baboon data set



```{r setup, echo = FALSE}
library(Cairo)

# global knitr options
knitr::opts_chunk$set(fig.path = 'figure/',
                      fig.prefix = testmethod,
                      dpi = 300,
                      dev = "png",
                      dev.args = list(type = "cairo-png"),
		                  fig.width = 10)
```


## Data characteristics


Full sample sizes by data set:

```{r samples, echo = FALSE, fig.width = 10, fig.height = 4 * length(containers)}
samples <- data.frame(Dataset=sapply(containers, mainExpName), N=sapply(containers, ncol)) %>% arrange(N)
kable(samples)
```

Feature counts by data set:

```{r chars, echo = FALSE, fig.width = 10, fig.height = 4*length(containers)}
feat <- df_benchmark %>% select(Dataset, Rank, Features) %>%
     	            group_by(Dataset, Rank) %>%
                    summarize(Features=max(Features)) %>% # Variation due to removing zero rows & cols per round
		    unique() %>%
                    pivot_wider(names_from=Dataset, values_from=Features, values_fill=0)
feat <- feat[order(feat[,2]), ] # Sort by the first dataset feature counts
kable(feat)
```


## Relative differences in execution time by sample size

```{r ratio1, echo = FALSE, fig.height = 5, fig.width = 4 * length(containers)}
df <- df_benchmark %>% select(Dataset, Time, Features, Samples, Rank, ObjectType) %>% 
      pivot_wider(names_from = c(ObjectType), values_from = Time) %>% 
      mutate(Ratio = tse / pseq) 

p <- ggplot(df, aes(x = Samples, y = Ratio, group = Rank, color = Features)) + 
    geom_point() + 
    geom_line() +
    labs(title = "Relative difference in execution times",
         x = "Samples (N)",
         y = "Relative time (TreeSE / phyloseq)",
         color = "Features (N)",
         caption = paste("Execution time ratio by samples (TreeSE/phyloseq): ", testmethod)) +
    scale_x_log10() +
    facet_grid(. ~ Dataset) + 
    geom_hline(yintercept = 1, linetype = 2) 

print(p)
```

```{r ratio2, echo = FALSE, fig.height = 5, fig.width = 4 * length(containers)}
if (length(tests[[testmethod]]) == 3) {
  
  df <- df %>% mutate(SpeedyRatio = tse / speedyseq)
  
  p <- ggplot(df, aes(x = Samples, y = SpeedyRatio, group = Rank, color = Features)) + 
    geom_point() + 
    geom_line() +
    labs(title = "Relative difference in execution times",
         x = "Samples (N)",
         y = "Relative time (TreeSE / speedyseq)",
         color = "Features (N)",
         caption = paste("Execution time ratio by samples (TreeSE/speedyseq): ", testmethod)) +
    scale_x_log10() +
    facet_grid(. ~ Dataset) + 
    geom_hline(yintercept = 1, linetype = 2) 

  print(p)

}
```


## Absolute execution time by sample size

```{r abs_by_time, echo = FALSE, fig.height = 13, fig.width = 3 * length(containers)}
df <- df_benchmark %>% select(Dataset, Time, Features, Samples, Rank, ObjectType)
df <- df %>% filter(Dataset %in% names(which(table(df$Dataset) >= 10))) %>%
             mutate(Rank = factor(Rank, levels=as.character(feat$Rank)))
cols <- c(tse = "black", pseq = "darkgray", speedyseq = "lightgray") # Ensure colors always in same order

p <- ggplot(df, aes(x = Samples, y = Time * 1000, color = ObjectType)) + 
    geom_point() + 
    geom_line() +
    theme(axis.text.x = element_text(angle = 90)) + 
    labs(title = "Absolute times",
         x = "Samples (N)",
         y = "Time (ms)",
         color = "Container",
         caption = paste("Execution time by sample size:", testmethod)) +
	 scale_color_manual(values = cols[unique(df$ObjectType)]) + 
    facet_grid(Rank ~ Dataset, scales = "free_y") 

print(p)
```


## Execution times vs number of features

```{r multi_ex_time, echo = FALSE, fig.height = 13, fig.width = 3 * length(containers)}
df <- df_benchmark 

p <- ggplot(df, aes(x = Features, y = Time * 1000, color = ObjectType)) + 
    geom_point() + 
    geom_line() +
    labs(title = "Feature count vs. execution time",
         x = "Features (D)",
         y = "Execution time (ms)",
         color = "Method:",
         caption = paste("Execution time by feature count:", testmethod)) +
    scale_x_log10() + # Log is often useful with sample size
    scale_color_manual(values = cols[unique(df$ObjectType)]) +     
    facet_grid(Samples ~ Dataset, scales = "free_y")

print(p)
```
